Dynamic Load Balancing for\\Heterogeneous Systems

### 2011-12-05 paper/paper.tex
Various typo changes.

\section{Load Balancing Benchmarks}
For this paper, three different benchmarks were used to assess the
effectiveness of the load balancer.  These benchmarks were chosen to represent
tasks that would actually be performed on a GPU, and have seen use in already
existing benchmark suites~\cite{Danalis2010}.  The applications that we have chosen
for this project are VectorAdd, VectorAdd+, and Reduce.

The VectorAdd application is a simple vector addition operation that computes the sum
of two input vectors, and stores the result into an output vector.  This represents the
vector operation $\vec{C} = \vec{A} + \vec{B}$.  This kernel is naive to parallelize, as
it exhibits embarrassingly parallel structure.  Each thread being run maps to a single 
index in the array and computes the output value for that index.  This algorithm is also
nicely designed for load-balancing because of the lack of data dependencies, reducing 
the amount of communication to the GPU.

The next benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
the VectorAdd benchmark, but it computes 1000 additions before storing the result
in the output vector.  The motivation behind this benchmark is to reduce memory
access overhead with more computation.  This benchmark was compiled without optimizations
so that the additions would not be optmized into a single multiplication.  The data
dependencies are exactly the same as for the VectorAdd application, so the results for
data transfer speeds should be equal for both applications, and the difference between
the two should only be the kernel execution times.

The Reduce application is an application that performs a reduction operation on a
large array of data.  In this case, the sum of a variably-sized array is computed.
The Reduce benchmark is unique in that different code is being run on the CPU and 
the GPU.  The reason for this is that the GPU cannot efficiently compute a reduction
the way a CPU typically would -- due to the relatively slower cores and much greater
memory latency.  A optimized reduction on the GPU is a multi-stage process, in which the
output of one step is used as the input of the next.  Each SM computes the reduction
of a smaller set of elements from the larger array during each iteration.  This
requires the use of numerous local atomic operations, as well as local memory to 
facilitate optimized memory access between threads on an SM.
This process continues until the reduction operation is only done by a single SM.
A natural consequence of this algorithm is that not all of the SMs will be utilized during
the computation.  In addition, inside an iteration all of the streaming processors
will only be used for the first stage of reduction, greatly reducing utilization on the
device.

All numbers reported in the results section
are given as a mean of ten runs for that particular benchmark, load-balancing scheme,
and data size.

@@

\section{Load Balancing Schemes}
A naive implementation of CPU+GPU co-processing will in many cases lead to an imbalance
of work for either the CPU or GPU.  To deal with this problem we will investigate the
use of 2 different load balancing schemes in addition to the performance of a CPU-only or
a GPU-only solution.  These schemes are illustrated in Figure~/ref{fig:scheduling}

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  Unfortunately, none of the static load balancing schemes were able
to outperform just the CPU; all the static times were somewhere in between the CPU-only and
GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
worse than all the other schemes, and more than 100\% worse than just the CPU.
100\% worse than just the CPU.

@@

\section{Conclusion}
It is clear that our original hypotheses were not supported by our results.  That
is, we were not able to see consistent, significant speedup from any of our load
balancing schemes.  In many scenarios, the GPU simply outperformed the CPU, with
the static load balancing schemes falling somewhere between the two, and the dynamic
load balancer being far worse than anything else.  There were only two notable
exceptions to this trend.  The first was the reduction benchmark on the discrete GPU
platform, where the CPU outperformed the GPU due to the low stream processor utilization
inherent in a GPU reduction implementation.  However, the static load balancing schemes
were still outperformed by the CPU in this case, so no overall speedup was gained. The
second exception was the vector add+ benchmark on the discrete GPU platform, where the
dynamic load balancer actually outperformed the tested static load balancing schemes.
However, this was due to the vast disparity in the GPU and CPU runtimes for this
particular benchmark; giving any significant amount of data to the CPU would hurt
performance.  With the dynamic load balancer, the CPU will only compute a few blocks
of data, rather than 25\% or more with the static load balancer.

@@

Based on the CPU-only and GPU-only runtimes of each benchmark, we were able to calculate an
optimal static data partitioning, such that the computations on the CPU and GPU took equal
time to complete.  Based on the disparity of runtimes between the CPU-only and GPU-only
results, the optimal runtimes ranged from a 2\% to nearly a 45\% speedup over just the GPU
alone.  Unfortunately, none of these theoretical results were supported in practice.  It seems
to be a limitation of OpenCL that it will not run kernels on different devices in parallel,
which effectively renders our load balancing schemes pointless.

Additionally, it seems that a fused CPU+GPU architecture may provide some performance benefits
over just a discrete GPU architecture.  Our results shows that the GPU outperformed the CPU on
the fused platform's reduction benchmark, but this was not the case for the discrete GPU platform.
However, as the CPUs and GPUs differed significantly between the two testing environments, we
cannot say definitively that one architecture will always be better than the other.

>>>>

\section{Load Balancing Benchmarks}
For this paper, three different benchmarks were used to assess the
effectiveness of the load balancer.  The first is a reduction benchmark,
in which the sum of variably-sized array is computed.  The reduction benchmark
is unique in that different code is being run on the CPU and the GPU.  The
reason for this is that the GPU cannot efficiently compute a reduction the way
a CPU typically would -- due to the relative inefficiency of backward branches
on the GPU.  A reduction on the GPU is a multi-stage process, in which the
output of one step is used as the input of the next.  Each steam processor
computes the reduction of two elements during each iteration, reducing the amount
of work during each step by a factor of two.  As a result, each step utilizes half
the stream processors of the previous step.  This continues until the reduction
operation is only done by a single stream processor.  A natural consequence of
this algorithm is that not all of the stream processors will be utilized during
the computation.

The next benchmark is a simple vector addition operation that computes the sum
of two input vectors, and stores the result into an output vector.  The third
benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
the Vector Add benchmark, but it computes 1000 additions before storing the result
in the output vector.  The motivation behind this benchmark is to reduce memory
access overhead with more computation.  All numbers reported in the results section
are given as a mean of ten runs for that particular benchmark, load-balancing scheme,
and data size.

@@

\section{Load Balancing Schemes}
A naive implementation of CPU+GPU co-processing will in many cases lead to an imbalance
of work for either the CPU or GPU.  To deal with this problem we will investigate the
use of 2 different load balancing schemes in addition to the performance of a CPU-only or
a GPU-only solution.  These schemes are illustrated in Figure~\ref{fig:scheduling}.

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  Unfortunately, none of the static load balancing schemes were able
to outperform just the CPU; all the static times were somewhere in between the CPU-only and
GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
worse than all the other schemes, and more than 100\% worse than just the CPU.

@@

\section{Conclusion}
It is clear that our original hypotheses were not supported by our results.  That
is, we were not able to see consistent, significant speedup from any of our load
balancing schemes.  In many scenarios, the GPU simply outperformed the CPU, with
the static load balancing schemes falling somewhere between the two, and the dynamic
load balancer being far worse than anything else.  There were only two notable
exceptions to this trend.  The first was the reduction benchmark on the discrete GPU
platform, where the CPU outperformed the GPU due to the low stream processor utilization
inherent in a GPU reduction implementation.  However, the static load balancing schemes
were still outperformed by the CPU in this case, so no overall speedup was gained. The
second exception was the vector add+ benchmark on the discrete GPU platform, where the
dynamic load balancer actually outperformed the tested static load balancing schemes.
However, this was due to the vast disparity in the GPU and CPU runtimes for this
particular benchmark, as giving any significant amount of data to the CPU would hurt
performance.  With the dynamic load balancer, the CPU will only compute a few blocks
of data, rather than 25\% or more with the static load balancer.

@@

Based on the CPU-only and GPU-only runtimes of each benchmark, we were able to calculate an
optimal static data partitioning, such that the computations on the CPU and GPU took equal
time to complete.  Based on the disparity of runtimes between the CPU-only and GPU-only
results, the optimal runtimes ranged from a 2\% to nearly a 45\% speedup over just the GPU
alone.  Unfortunately, none of these theoretical results were supported in practice.  It seems
to be a limitation of OpenCL that it will not run kernels on different devices in parallel,
which effectively renders our current load balancing schemes pointless.

Additionally, it seems that a fused CPU+GPU architecture may provide some performance benefits
over just a discrete GPU architecture.  Our results show that the GPU outperformed the CPU on
the fused platform's reduction benchmark, but this was not the case for the discrete GPU platform.
However, as the CPUs and GPUs differed significantly between the two testing environments, we
cannot say definitively that one architecture will always be better than the other.
